Complementary roles of Theta and Alpha in Multi-Item Associations

### 2019-07-15 manuscript.tex
revisions from randy

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declaritive memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically, oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding and retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the information that is encoded into the underlying representation.

@@

This accumulated evidence suggests a relationship between theta and alpha frequency bands and long-term memory systems as constructed by the Complimentary Learning Systems theory \cite{McClellandMcNaughtonOReilly95,OReillyBhattacharyyaHowardEtAl14}.  Namely, that theta power positively tracks successful engagement (encoding/retrieval) between hippocampus and neocortex, while alpha power negatively tracks with this process as active inhibition of cortex is released in response to the exchange of information between hippocampus and neocortex.  This relationship is further supported by the connectivity of the thalamus with cortex such that a specific model for how information is routed into and out of these systems can be tied to empirical measures of neural oscillations.  This work seeks to test this relationship through a series of computational simulations of hippocampal-thalamo-cortical interaction, and an accompanying EEG study in humans to compare against simulations results.

@@

\subsection{Electrophysiological recordings}
A 128-channel HydroCel Geodesic Sensor NetTM (GSN 200, v. 2.1; Tucker, 1993) was used to measure the EEG at the scalp using a central vertex reference (Cz) with a sampling rate of 250 Hz and a low-pass hardware filter at 100 Hz. The net was connected to a DC-coupled, high-input impedance amplifier (200 M$\Omega$, Net AmpsTM; Electrical Geodesics, Inc., Eugene, OR) and recordings were made using the Net Station application. The electrodes were adjusted until impedance measurements were less than 40 k$\Omega$.

@@

\subsection{Representation Similarity Analysis}
In an attempt to test for non-target reactivation a Representation Similarity Analysis was done across retrieval attempts within a given event.  The interpretation of this analysis is that events with more non-target reactivation across the various cued recognition retrieval attempts would show more similarity in their EEG time frequency topography.  The first approach, see Section \ref{sec:deevSpecAnal} for a more complete description, tried to maintain as much alignment with standard RSA as possible, by doing correlations using the full frequency spectrum (3 to 50 Hz), and the full time window (500 ms to 3000 ms for Stimulus Locked epochs, and -3000 ms to -500 ms Response Locked).  Averaging pairwise correlations across all retrieval attempts within a given event yielded a single measure of similarity for each event, which were then averaged into open and closed-loop conditions. No significant differences were found between open and closed-loop events for both stimulus locked and response locked epochs.  It is likely the full $time \times frequency \times electrode$ space is non-optimal for detecting this similarity differences, and in an attempt to narrow down this space a second approach was adopted to take advantage of the clustering methods used in the power analysis.

@@

What are the mechanisms in the model which support the differences witnessed between open and closed-loop association structures? The main cause of these condition differences likely stem from the model's ability to reactivate the non-target elements more so in the closed-loop compared to the open-loop condition.  This increased reactivation initially comes from the fact that each of the different event elements are studied together in the closed-loop events, allowing the network to be more likely to later reactivate those associations.  In contrast, because the open-loop events don't have the fully overlapping association structure that the close loop events do, a given cue in the open-loop retrieval trial is only likely to reactive its previously studied association which would be the target element of a given retrieval trial, and unable to reactivate the non-target element.

>>>>

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), and alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declaritive memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically, oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding and retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the information that is encoded into the underlying representation.

@@

This accumulated evidence suggests a relationship between theta and alpha frequency bands and long-term memory systems as constructed by the Complimentary Learning Systems theory \cite{McClellandMcNaughtonOReilly95,OReillyBhattacharyyaHowardEtAl14}.  Namely, that theta power positively tracks successful engagement (encoding/retrieval) between hippocampus and neocortex, while alpha power negatively tracks with this process as active inhibition of cortex is released in response to the exchange of information between hippocampus and neocortex.  This relationship is further supported by the connectivity of the thalamus with cortex such that a specific model for how information is routed into and out of these systems can be tied to empirical measures of neural oscillations.  This work seeks to test this relationship through a series of computational simulations of hippocampal-thalamo-cortical interaction, and an accompanying EEG study in humans to compare against simulation results.

In general, our results show that our model can capture broad time frequency dynamics related to alpha and theta oscillations, and that it can be used to evaluate patterns of human EEG against an underlying theoretical model.  This model provides explanatory power in understanding how thamalmo-cortical attentional mechanisms can enhance the binding between multi-item associations through reactivation, as well as diminish those associations when the underlying encoding structure does not promote reactivation.

@@

\subsection{Electrophysiological recordings}
A 128-channel HydroCel Geodesic Sensor Net (GSN 200, v. 2.1; Tucker, 1993) was used to measure the EEG at the scalp using a central vertex reference (Cz) with a sampling rate of 250 Hz and a low-pass hardware filter at 100 Hz. The net was connected to a DC-coupled, high-input impedance amplifier (200 M$\Omega$, Net AmpsTM; Electrical Geodesics, Inc., Eugene, OR) and recordings were made using the Net Station application. The electrodes were adjusted until impedance measurements were less than 40 k$\Omega$.

@@

\subsection{Representation Similarity Analysis}
In an attempt to test for non-target reactivation a Representation Similarity Analysis was done across retrieval attempts within a given event.  The interpretation of this analysis is that events with more non-target reactivation across the various cued recognition retrieval attempts would show more similarity in their EEG time frequency topography.  The first approach, see Section \ref{sec:deevSpecAnal} for a more complete description, tried to maintain as much alignment with standard RSA as possible, by doing correlations using the full frequency spectrum (3 to 50 Hz), and the full time window (500 ms to 3000 ms for Stimulus Locked epochs, and -3000 ms to -500 ms Response Locked).  Averaging pairwise correlations across all retrieval attempts within a given event yielded a single measure of similarity for each event, which were then averaged into open and closed-loop conditions. No significant differences were found between open and closed-loop events for both stimulus locked and response locked epochs.  It is likely the full $time \times frequency \times electrode$ space is non-optimal for detecting these similarity differences, and in an attempt to narrow down this space a second approach was adopted to take advantage of the clustering methods used in the power analysis.

@@

What are the mechanisms in the model which support the differences witnessed between open and closed-loop association structures? The main cause of these condition differences likely stem from the model's ability to reactivate the non-target elements more so in the closed-loop compared to the open-loop condition.  This increased reactivation initially comes from the fact that each of the different event elements are studied together in the closed-loop events, allowing the network to be more likely to later reactivate those associations.  In contrast, because the open-loop events don't have the fully overlapping association structure that the closed loop events do, a given cue in the open-loop retrieval trial is only likely to reactivate its previously studied association which would be the target element of a given retrieval trial, and unable to reactivate the non-target element.
